Biogeography of crop progenitors and wild plant resources during the transition to agriculture in West Asia, 21–8.3 ka

### 2025-01-03 analysis/paper.qmd
Refactor as manuscript

---
title: Biogeography of crop progenitors and wild plant resources during the transition to agriculture in West Asia
date: today
author: 
- name: Joe Roe
  id: jr
  orcid: 0000-0002-1011-1244
  email: joeroe@hey.com
  affiliation: University of Bern
- name: Amaia Arranz-Otaegui
  id: aao
  orcid: 0000-0002-5091-6426
  affiliation: University of the Basque Country
abstract: |
  This paper presents the first comprehensive reconstructions of the prehistoric 
  ranges of [X] plant species known to have been gathered or cultivated by early
  agricultural societies in Southwest Asia, including the progenitors of the 
  first crops. We used machine learning to train an ecological niche model (ENM)
  of each species based on its present-day distribution in relation to climate 
  and environmental variables. Predictions of the potential ranges of these 
  species at key stages of the Pleistocene–Holocene transition could then be
  derived from these models using palaeoclimate 
  Species ranges were on average [X%] [larger|smaller] in the Early Holocene 
  compared to today, indicating [...]. The modelled ranges predict the
  observed occurrence of these species on archaeological sites with [low|medium|high]
  accuracy. The regional ubiquity of species in the archaeological record is [not] 
  correlated with the predicted size of its range and the diversity of
  archaeobotanical assemblages is [not] correlated with the predicted diversity
  of its environs. This indicates that trends in taxonomic composition of
  the archaeobotanical record is [not] likely to be influenced by environmental 
  change and species turnover, not just, as is often assumed, human economic 
  choices.
format: html
bibliography: references.bib
---

```{r dependencies}
library("cowplot")
library("dplyr")
library("dotenv")
library("forcats")
library("fs")
library("furrr")
library("ggplot2")
library("ggrepel")
library("ggspatial")
library("gt")
library("here")
library("khroma")
library("progressr")
library("purrr")
library("ranger")
library("readr")
library("recipes")
library("rgbif")
library("scales")
library("sf")
library("stars")
library("stringr")
library("tibble")
library("tidymodels")
library("tidyr")

@@

```{r define-periods}
archaeo_periods <- tribble(
  ~period,             ~agriculture,                          ~start_bp, ~end_bp,
  "Late Epipal.",      "Foraging",                            15000,     11700,
  "PPNA",              "Pre-domestication cultivation",       11700,     10700,
  "EPPNB",             "Cultivation of domesticated species", 10700,     10200,
  "MPPNB",             "Cultivation of domesticated species", 10200,     9500,
  "LPPNB/C",           "Agriculture",                         9500,      8500,
  "Pottery Neolithic", "Agriculture",                         8500,      6500,
  "Chalcolithic",      "Agriculture",                         6500,      5000,
)

@@

## Introduction

* The first farming societies had an ecological context
* Subsistence is understood (largely) through archaeobotany; ecological context from archaeobotany, palynology, palaeoclimate, etc.
  * But these have a variety of biases (human selection, taphonomy, etc.)
  * And at the end of the day only represent specific places – interpolation to the entire region is not straightforward
  * Overlap makes it difficult to see where human choices depart from the environmental background [cf. Martin et al. 2016, for zooarch]
* Here we present an alternative approach using ENM
  * Whole region, at multiple climate snapshots
  * Independent of archaeobot. and pal. clim. data, so can verify and compare

## Background

* Epipal./Neo. plant-based economies
  * More than the "founder crops"; 
  * More than food;
  * Regionally and temporally diverse
  * ...so we model lots of species!
* Regional ecological reconstructions
  * See CSEAS (AEA-prep) presentation
  * Figure: comparisons
* Archaeological applications of ENM
  * Review: Franklin et al. 2015
  * In SW Asia: Conolly et al. 2012, Collins et al. 2018
  * Train with archaeo, or train with present-day?
    * In ENM literature, many sceptical about fore- and hind-casting
    * Archaeo data a golden opportunity here

## Methods and materials

@@

### Predictor data

Bioclimatic data – the standard, and what allows us to do hindcasting.

Topographic data derived from a DEM:

* Aspect and slope – proxies for solar radiation, but only if high resolution [@AustinVanNiel2011; @LeempoelEtAl2015]
* TWI – proxy for soil moisture, important for arid environments [@KopeckyCizkova2010; @CamposEtAl2016; DiVirgilioEtAl2018]

Soil data – important for plants [@DubuisEtAl2013; @ModEtAl2016; @VelazcoEtAl2017] but not all variables from SoilGrids are reliable: so we just use soil texture (clay, silt, sand) and pH at the surface (0-5 cm depth) [@MillerEtAl2024].

@@

@@

### Occurrence data

```{r read-archaeo_flora}
archaeo_flora <- read_tsv(raw_data("swasia_neolithic_flora.tsv"))
```

```{r tidy-archaeo_flora}
archaeo_flora |>
  rename(taxon_group = "taxon", taxon = "taxon_detail") |>
  # Assign to archaeological periods
  # TODO: refactor as function
  # TODO: assign climate periods too
  mutate(
    age_mid = age_end + ((age_start - age_end) / 2),
    period = cut(-age_mid, 
                 breaks = -c(archaeo_periods$start_bp, 0), 
                 labels = archaeo_periods$period,
                 ordered_result = TRUE)
  ) ->
  archaeo_flora
```

```{r summarise-archaeo_flora}
n_assemb_neolithic <- archaeo_flora |> 
  distinct(site_name, phase_code, period) |> 
  filter(period %in% neolithic) |> 
  nrow()

n_site_neolithic <- archaeo_flora |> 
  filter(period %in% neolithic) |> 
  distinct(site_name) |> 
  nrow()
```

To generate the list of taxa modelled in this study, we ...

* Taxa present at more than three Neolithic sites, according to the database of our previous study [@ArranzOtaeguiRoe2023]
* Identified at least to species (and matchable to the GBIF Backbone Taxonomy)
* Have at least 100 occurrences in West Asia (meeting the criteria below) in the GBIF database
* Excluded domestic species, replacing them with their wild progenitor, where known and present in the GBIF database

We also considered the number modern occurrences available for each taxon.
ENMs are known to have reasonable predictive power even with small training samples (i.e. less than a hundred observations, [@StockwellPeterson2002; @HernandezEtAl2006; @WiszEtAl2008]).

```{r generate-flora}
archaeo_flora |>
  filter(period %in% neolithic) |>
  group_by(taxon) |>
  summarise(
    n_present = length(unique(phase_code)),
    p_present = n_present / n_assemb_neolithic,
    .groups = "drop"
  ) |>
  filter(n_present > 2) ->
  flora
```

```{r tidy-flora-taxonomy}
# Normalise species names and match to the GBIF Backbone Taxonomy
# Excluding taxa not identified to species
flora |>
  filter(!str_ends(taxon, "sp.")) |>
  mutate(
    taxon = case_match(taxon,
                       "Vicia narbonense" ~ "Vicia narbonensis", # N=3
                       "Triticum aestivocompactum" ~ "Triticum aestivum compactum", # N=3
                       "Heliotropium persicum" ~ "Moltkia angustifolia", # N=2
                       "Lens orientalis" ~ "Vicia orientalis", # N=2
                       "Scirpus tabernaemontani" ~ "Schoenoplectus tabernaemontani", # N=2
                       "Stipa capensis" ~ "Stipa dregeana", # N=2
                       .default = taxon),
    taxon_original = taxon,
    taxon = map_chr(taxon, gbif_accepted_name)
  ) ->
  flora
```

```{r tidy-flora-domesticates}
# Recode domesticates as their wild progenitors.
# Add other wild progenitors of wheats
flora |> 
  mutate(taxon = case_match(
    taxon,
    "Cicer arietinum" ~ "Cicer reticulatum",
    "Hordeum vulgare" ~ "Hordeum spontaneum",
    "Vicia lens" ~ "Vicia orientalis",
    "Linum usitatissimum" ~ "Linum bienne",
    "Triticum aestivum" ~ "Triticum turgidum dicoccum",
    "Triticum dicoccoides" ~ "Triticum turgidum dicoccum",
    "Triticum turgidum durum" ~ "Triticum turgidum dicoccum",
    "Triticum monococcum" ~ "Triticum monococcum aegilopoides",
    .default = taxon
  )) |>
  bind_rows(tibble(taxon = c("Triticum urartu", "Aegilops speltoides", 
                             "Aegilops tauschii"))) ->
  flora
```

```{r fetch-occ-count}
flora |>
  group_by(taxon) |>
  summarise(
    taxon_original = list(c(taxon_original)),
    n_present = max(n_present), # TODO: be more precise?
    p_present = max(p_present)
    ) |>
  mutate(n_occ = map_dbl(taxon, \(x) occ_count(scientificName = x, 
                                               hasCoordinate = TRUE, 
                                               hasGeospatialIssue = FALSE,
                                               geometry = bbox_wkt(w_asia)))) ->
  flora
```

```{r filter-flora}
flora <- filter(flora, n_occ > 50)
# TODO: Temporary measure.
# Why are there 47k occurrences of Avena sterilis??
flora <- filter(flora, taxon != "Avena sterilis")
```

Occurrence data was obtained from GBIF using the GBIF occurrence API.
This deals with synonyms etc. (using the GBIF Backbone Taxonomy as a standard), so e.g. retrieving data for "Bolboschoenus" includes all species and subspecies, including specimens described as *Bolboschoenus sp.*, *Bolboschoenus maritimus* and *Scirpus maritimus*.

<!-- TODO: following @ValaviEtAl2022, more accurate to call pseudo-absences
  "background samples" -->
Occurrence data only tells us where a species is present;
there is no definitive information on where the species is *not* found.
We therefore need to generate random background points or "pseudo-absences" to feed to the model.
There are several ways to do this. 
We follow the advice of Barbet-Massin et al. (2012) for regression-based species distribution models and use a large (~10000) random sample of points, weighted equally against the presences in the regression.
@ValaviEtAl2022 recommends using a very large background sample for random forest models.

@@

TODO: Soil, terrain (incl. TWI, slope?)

### Random Forest

There are many methods for ENM.
Random Forest is one of the best performing, providing it is appropriately parameterised to account for the class imbalance between presence and background samples (balanced random forest) [@ValaviEtAl2021a; @ValaviEtAl2021b].

We used the R package ranger [@WrightZiegler2017].

```{r prepare-niche-model-data}
# Split training, test (for model assessment) and cross-validation sets (for
# tuning)
flora <- flora |>
  mutate(
    occ = map(occ, \(x) mutate(x, weight = importance_weights(weight))),
    occ = map(occ, \(x) drop_na(x, starts_with("bio"))), # TODO: where do these come from?
    occ_split = map(occ, \(x) initial_split(x, strata = present)),
    # occ_folds = map(occ_split, \(x) vfold_cv(training(x), v = 10, strata = present))
  )

@@

```{r niche-models}
# Random Forest Niche Model
# With hyperparameters adapted from the example of a down-sampled RF model given 
# by https://rvalavi.github.io/SDMwithRFs/#model-fitting-with-ranger-package
niche_model <- rand_forest(mode = "classification", trees = 1000) |>
  set_engine(
    "ranger", 
    sample.fraction = c(sum(as.logical(.y())) / n_background, 1),
    probability = TRUE
  )

@@

niche_workflow <- workflow(niche_recipe, niche_model)
```

We follow @ValaviEtAl2021a in using down-sampling and their recommended hyperparameters to avoid overfitting.

@@

```{r tune-niche-models}
# Tune model (separately for each taxon)
# TODO: would be nice but is extremely slow
# flora |>
#   slice_sample(n = 10) |>
#   mutate(
#     niche_tuning_grid = map(occ_split, function(data, recipe) {
#       grid_regular(
#         finalize(mtry(), bake(prep(recipe), training(data), all_predictors())),
#         trees(),
#         min_n(),
#         levels = 5
#       )
#     }, recipe = niche_recipe, .progress = TRUE),
#     niche_tuning = map2(niche_tuning_grid, occ_folds, function(grid, folds, wflow) {
#       tune_grid(wflow, folds, grid = grid)
#     }, wflow = niche_workflow, .progress = TRUE),
#     niche_hparams = map(niche_tuning, select_best, metric = "roc_auc"),
#     niche_model = map(niche_hparams, function(params, wflow) {
#       finalize_workflow(wflow, params)
#     }, wflow = niche_workflow)
#   )
```

@@

The output of the model is probabilistic.
However, this should not be understood as an actual probability of occurrence [@CITE], but more akin to an estimate of habitat suitability.
To simplify interpretation, we can convert these predictions into binary presence/absence maps, a process called "thresholding".
We select the threshold value for each model individually, using MaxSSS (as recommended by [@LuiEtAl2013])
This also makes it possible to analyses the predictions together as an assemblage.

@@

## Results

### Model assessment

@@

### Modelled ecological niches

* On current data
* Sensitivity to climate fluctuations

### Hindcasting

@@

### Comparison to archaeological occurrences

## Discussion

@@

@@

## Acknowledgments {.appendix}

>>>>

---
title: Biogeography of crop progenitors and wild plant resources during the transition to agriculture in West Asia, 21–8.3 ka
date: today
author: 
- name: Joe Roe
  id: jr
  orcid: 0000-0002-1011-1244
  email: joeroe@hey.com
  affiliations:
    - name: University of Bern
      country: Switzerland
    - name: University of Copenhagen
      country: Denmark
  corresponding: true
- name: Amaia Arranz-Otaegui
  id: aao
  orcid: 0000-0002-5091-6426
  affiliation: 
    - name: University of the Basque Country
      country: Spain
abstract: |
  This paper presents the first comprehensive reconstructions of the 
  palaeodistributions of [X] plant species known to have been gathered or 
  cultivated by early agricultural societies in Southwest Asia, including the 
  progenitors of the first crops. We used machine learning to train an 
  ecological niche model (ENM) of each species based on its present-day 
  distribution in relation to climate and environmental variables. Predictions 
  of the potential ranges of these species at key stages of the 
  Pleistocene–Holocene transition could then be derived from these models using 
  hindcast data from palaeoclimate simulations.
  Species ranges were on average [X%] [larger|smaller] in the Early Holocene 
  compared to today, indicating [...]. The modelled ranges predict the
  observed occurrence of these species on archaeological sites with [low|medium|high]
  accuracy. The regional ubiquity of species in the archaeological record is [not] 
  correlated with the predicted size of its range and the diversity of
  archaeobotanical assemblages is [not] correlated with the predicted diversity
  of its environs. This indicates that trends in taxonomic composition of
  the archaeobotanical record is [not] likely to be influenced by environmental 
  change and species turnover, not just, as is often assumed, human economic 
  choices.
format: 
  elsevier-pdf:
    keep-tex: true
    cite-method: citeproc
    journal:
      name: Open Quaternary
      formatting: review
bibliography: references.bib
execute: 
  echo: false
  eval: false
---

* [ ] Finish analysis:
  - [ ] Add terrain, soil
  - [ ] Find a resolution that's workable
  - [ ] Do thresholding?
- [ ] First draft:
  - [ ] Introduction
  - [ ] Background
    - [ ] Biogeography
    - [ ] ENM
  - [x] Methods & Materials
    - [x] Occurrence data
    - [x] Predictor data
    - [x] Random Forest model
  - [ ] Results
    - [ ] Model assessment
    - [ ] Hindcasting
  - [ ] Discussion
    - [ ] General trends
    - [ ] Case studies
  - [ ] Conclusion
- [ ] Figures
  - [ ] tbl-predictors
  - [ ] tbl-climate-periods
* [ ] Appendix with all hindcast predictions
- [ ] References & copyediting
- [ ] Final proofread

```{r dependencies, eval=TRUE, message=FALSE}
library("cowplot")
library("dplyr")
library("dotenv")
library("english")
library("forcats")
library("fs")
library("furrr")
library("ggplot2")
library("ggrepel")
library("ggspatial")
library("gt")
library("here")
library("khroma")
library("progressr")
library("purrr")
library("ranger")
library("readr")
library("recipes")
library("rgbif")
library("scales")
library("sf")
library("stars")
library("stringr")
library("tibble")
library("tidymodels")
library("tidyr")

@@

```{r define-periods, eval=TRUE}
archaeo_periods <- tribble(
  ~period,             ~agriculture,                          ~start_bp, ~end_bp,
  "Late Epipal.",      "Foraging",                            15000,     11700,
  "PPNA",              "Pre-domestication cultivation",       11700,     10700,
  "EPPNB",             "Cultivation of domesticated species", 10700,     10200,
  "MPPNB",             "Cultivation of domesticated species", 10200,     9500,
  "LPPNB/C",           "Agriculture",                         9500,      8500,
  "Pottery Neolithic", "Agriculture",                         8500,      6500,
  "Chalcolithic",      "Agriculture",                         6500,      5000,
)

@@

# Introduction

* The first farming societies had an ecological context
* Subsistence is understood (largely) through archaeobotany and zooarchaeology; ecological context from environmental archaeology, palynology, palaeoclimate records, etc.
  * But these have a variety of biases (human selection, taphonomy, etc.)
  * And at the end of the day only represent specific places – interpolation to the entire region is not straightforward
  * Overlap makes it difficult to see where human choices depart from the environmental background [cf. @MartinEtAl2016]
  * Or to fully contextualise subsistence strategies and shifts in strategies in response to environmental shifts [e.g. @YaworskyEtAl2023].
* Here we present an alternative approach using ENM
  * Whole region, at multiple climate snapshots
  * Independent of archaeobot. and pal. clim. data, so can verify and compare

# Background

* The transition to agriculture in West Asia was...

## Biogeography and agricultural origins

* Has always been important in study of agricultural origins
  * Historically: Vavilov, Pumpelly & Childe
  * Genetic studies tell us origin points, but not ranges
* Epipal./Neo. plant-based economies were diverse
  * More than the "founder crops"; 
  * More than food
  * (In archaebot., not all intentionally collected)
  * Regionally and temporally diverse
  * ...so we model lots of species!
* Regional ecological reconstructions generally rely on the 'expert interpolation' (or what do they call it with isoscapes?) method
  * See CSEAS (AEA-prep) presentation
  * Figure: comparisons

## Ecological niche modelling in archaeology

* Review: @FranklinEtAl2015
* In SW Asia: @ConollyEtAl2012 (first use ever??), @CollinsEtAl2018
* Varying uses: to reconstruct human niches (esp. in Palaeolithic), to reconstruct spread of (esp. domestic) species from archaeological record, or to reconstruct wild ranges from modern occurrences and palaeoclimate (our approach)
* Caveats
  * In ENM literature, many sceptical about fore- and hind-casting
  * Archaeo data a golden opportunity here

# Methods and materials

@@

## Occurrence data

```{r data-flora}
# Flora present at archaeological sites
archaeo_flora <- read_tsv(raw_data("swasia_neolithic_flora.tsv")) |>
  rename(taxon_group = "taxon", taxon = "taxon_detail") |>
  # Assign to archaeological periods
  # TODO: refactor as function
  # TODO: assign climate periods too
  mutate(
    age_mid = age_end + ((age_start - age_end) / 2),
    period = cut(-age_mid, 
                 breaks = -c(archaeo_periods$start_bp, 0), 
                 labels = archaeo_periods$period,
                 ordered_result = TRUE)
  ) ->
  archaeo_flora

n_assemb_neolithic <- archaeo_flora |> 
  distinct(site_name, phase_code, period) |> 
  filter(period %in% neolithic) |> 
  nrow()

n_site_neolithic <- archaeo_flora |> 
  filter(period %in% neolithic) |> 
  distinct(site_name) |> 
  nrow()

# Flora to be modelled
archaeo_flora |>
  filter(period %in% neolithic) |>
  group_by(taxon) |>
  summarise(
    n_present = length(unique(phase_code)),
    p_present = n_present / n_assemb_neolithic,
    .groups = "drop"
  ) |>
  filter(n_present > 2) |>
  # Exclude taxa not identified to species
  filter(!str_ends(taxon, "sp.")) ->
  flora
```

We consider X<!-- TODO --> distinct taxa (@tbl-occ-count) - all the identifiable species known to be present at more than three Neolithic sites in West Asia, according to our previous study [@ArranzOtaeguiRoe2023].
<!-- TODO: something about not distinguishing between source of plant remains? -->

```{r data-flora-taxonomy}
# - Normalise species names and match to the GBIF Backbone Taxonomy
# - Recode domesticates as their wild progenitors.
# - Add other wild progenitors of wheats
flora <- flora |>
  mutate(
    taxon = case_match(taxon,
                       "Vicia narbonense" ~ "Vicia narbonensis", # N=3
                       "Triticum aestivocompactum" ~ "Triticum aestivum compactum", # N=3
                       "Heliotropium persicum" ~ "Moltkia angustifolia", # N=2
                       "Lens orientalis" ~ "Vicia orientalis", # N=2
                       "Scirpus tabernaemontani" ~ "Schoenoplectus tabernaemontani", # N=2
                       "Stipa capensis" ~ "Stipa dregeana", # N=2
                       .default = taxon),
    taxon_original = taxon,
    taxon = map_chr(taxon, gbif_accepted_name),
    taxon = case_match(
      taxon,
      "Cicer arietinum" ~ "Cicer reticulatum",
      "Hordeum vulgare" ~ "Hordeum spontaneum",
      "Vicia lens" ~ "Vicia orientalis",
      "Linum usitatissimum" ~ "Linum bienne",
      "Triticum aestivum" ~ "Triticum turgidum dicoccum",
      "Triticum dicoccoides" ~ "Triticum turgidum dicoccum",
      "Triticum turgidum durum" ~ "Triticum turgidum dicoccum",
      "Triticum monococcum" ~ "Triticum monococcum aegilopoides",
      .default = taxon
    )
  ) |>
  bind_rows(tibble(taxon = c("Triticum urartu", "Aegilops speltoides", 
                             "Aegilops tauschii")))
```

Taxonomic names were resolves to the canonical form specified in the GBIF Backbone Taxonomy [@GBIFSecretariat2023].
<!-- TODO: still relevent? aren't we modelling species? -->So for example occurrences for "Bolboschoenus" included all species and subspecies, including specimens described as *Bolboschoenus sp.*, *Bolboschoenus maritimus* and *Scirpus maritimus*.
Domestic species meeting our inclusion criteria were substituted with their wild progenitor(s), where known.

## Occurrence data

```{r data-occ-count}
flora |>
  group_by(taxon) |>
  summarise(
    taxon_original = list(c(taxon_original)),
    n_present = max(n_present), # TODO: be more precise?
    p_present = max(p_present)
    ) |>
  mutate(n_occ = map_dbl(taxon, \(x) occ_count(scientificName = x, 
                                               hasCoordinate = TRUE, 
                                               hasGeospatialIssue = FALSE,
                                               geometry = bbox_wkt(w_asia)))) ->
  flora
```

```{r filter-flora}
flora <- filter(flora, n_occ > 50)
# Why are there 47k occurrences of Avena sterilis??
flora <- filter(flora, taxon != "Avena sterilis")
```

Occurrence data was obtained from the Global Biodiversity Information Facility (GBIF) using via its application programming interface and the R package rgbif [@ChamberlainBoettiger2017; @rgbif].
Although ENMs have reasonable predictive power even with small training samples [@StockwellPeterson2002; @HernandezEtAl2006; @WiszEtAl2008], we excluded X<!-- TODO --> taxa with less than fifty recorded occurrences in our study region.
We also excluded one taxon (*Avena sterilis*) with over 47,000 occurrences, as this would have been computationally prohibitive and we were uncertain what account for such a disproportionately high number of records.

<!-- TODO: occurrence tidying -->

<!-- TODO: fix informal tone -->
Occurrence data only tells us where a species is present;
there is rarely definitive information on where the species is *not* found.
We therefore need to generate random background points or "pseudo-absences" to feed to the model.
There are several ways to do this. 
We follow the advice of @BarbetMassinEtAl2012 for regression-based species distribution models and use a large (:10000) random sample of points, weighted equally against the presences in the regression.
@ValaviEtAl2022 also recommend using a very large background sample for random forest models.

## Predictor data

We modelled the occurrence of species as a function of <!-- TODO -->X spatial predictor variables (@tbl-predictors).
These included:

@@

* Sixteen 'bioclimatic' variables derived from monthly temperature and precipitation values, following standard practice for species distribution models [@HijmansEtAl2005]. Contemporary bioclimatic predictor data for West Asia was extracted from the global CHELSA dataset [@KargerEtAl2017], which predicts temperature and precipitation from downscaled general circulation model output at 1 km resolution.

* Terrain aspect and slope, which at high resolution perform well as proxies for solar radiation when modelling plant occurrence [@AustinVanNiel2011; @LeempoelEtAl2015]; and the topographic wetness index (TWI), which serves as a proxy for soil moisture and is particularly important in modelling arid environments [@KopeckyCizkova2010; @CamposEtAl2016; @DiVirgilioEtAl2018]. All three were derived from the SRTM15+ digital elevation model using algorithms from WhiteboxTools [@Lindsay2016].

* Edaphic data from SoilGrids [@HenglEtAl2014; @HenglEtAl2017], which improves model performance for plants [@DubuisEtAl2013; @ModEtAl2016; @VelazcoEtAl2017]. Based on a recent assessment of the reliability of SoilGrids data for species distribution modelling [@MillerEtAl2024], we used a subset of four variables relating to soil texture (clay, silt, sand) and pH at the surface (0-5 cm depth).

@@

Predictor data was transformed to the same projection system (WGS84 / UTM 37 N) and a common resolution of X<!-- TODO --> km.

For hindcasting, we used reconstructed bioclimatic data for `r english(nrow(climate_periods) - 1)` key periods (@tbl-climate-periods) generated from downscaled paleoclimate simulations from the HadCM3 general circulation model [@BrownEtAl2018].
Terrain and soil predictors were held constant, since reconstructions of these variables in the past are not available at sufficient scale.
It is not likely that either macroscale topography or soil characteristics have altered significantly over the period of time considered here, so we assume that this does not degrade model performance, and may in fact benefit it by providing 'anchoring' predictors that are independent of climate change.

@@

## Random Forest

```{r niche-data}
# Split training, test (for model assessment) and cross-validation sets (for
# tuning)
flora <- flora |>
  mutate(
    occ = map(occ, \(x) mutate(x, weight = importance_weights(weight))),
    occ = map(occ, \(x) drop_na(x, starts_with("bio"))), # TODO: where do these come from?
    occ_split = map(occ, \(x) initial_split(x, strata = present)),
    # occ_folds = map(occ_split, \(x) vfold_cv(training(x), v = 10, strata = present))
  )

@@

Ecological niche modelling is a classification problem that can be approached with a wide range of statistical methods.
A substantial literature exists on the relatively performance of these approaches and their respective parameterisations [reviewed in @ValaviEtAl2022].
Random Forest, a widely-used machine learning algorithm, is amongst the best performing approaches for presence-only species distribution models, providing it is appropriately parameterised to account for the class imbalance between presence and background samples [@ValaviEtAl2021; @ValaviEtAl2022].
For our application, it also has the advantage of requiring little to no manual parameter tuning to achieve good predictive results, which makes it easier to model a larger numbers of taxa.

```{r niche-model}
# Random Forest Niche Model
# With hyperparameters adapted from the example of a down-sampled RF model given 
# by https://rvalavi.github.io/SDMwithRFs/#model-fitting-with-ranger-package
niche_model <- rand_forest(mode = "classification", trees = 1000) |>
  set_engine(
    "ranger", 
    sample.fraction = c(sum(as.logical(.y())) / n_background, 1),
    probability = TRUE
  )

@@

niche_workflow <- workflow(niche_recipe, niche_model)

# Tune model (separately for each taxon)
# TODO: questioning - would be nice but is extremely slow
# flora |>
#   slice_sample(n = 10) |>
#   mutate(
#     niche_tuning_grid = map(occ_split, function(data, recipe) {
#       grid_regular(
#         finalize(mtry(), bake(prep(recipe), training(data), all_predictors())),
#         trees(),
#         min_n(),
#         levels = 5
#       )
#     }, recipe = niche_recipe, .progress = TRUE),
#     niche_tuning = map2(niche_tuning_grid, occ_folds, function(grid, folds, wflow) {
#       tune_grid(wflow, folds, grid = grid)
#     }, wflow = niche_workflow, .progress = TRUE),
#     niche_hparams = map(niche_tuning, select_best, metric = "roc_auc"),
#     niche_model = map(niche_hparams, function(params, wflow) {
#       finalize_workflow(wflow, params)
#     }, wflow = niche_workflow)
#   )
```

For each taxon we trained a classification model to predict occurrence (presence or absence/background) based on our X<!--TODO--> predictor variables (@tbl-predictors).
We used the Random Forest algorithm implemented in the R package 'ranger' [@WrightZiegler2017] and the 'tidymodels' [@tidymodels] framework for data preprocessing and model selection.
To avoid overfitting, we follow @ValaviEtAl2021 in their recommended hyperparameters and use of down-sampling to balance presence and background samples.
Models for each taxon were fit independently, with redundant zero-variance predictors excluded, and assessed based on balanced training (3/4) and test (1/4) partitions.

@@

@@

<!-- TODO: cut? -->
The output of the model is probabilistic.
However, this should not be understood as an actual probability of occurrence [@CITE], but more akin to an estimate of habitat suitability.
To simplify interpretation, we can convert these predictions into binary presence/absence maps, a process called "thresholding".
We select the threshold value for each model individually, using MaxSSS [as recommended by @LiuEtAl2013].
This also makes it possible to analyses the predictions together as an assemblage.

@@

# Results

## Model assessment

@@

* Modelled ecological niches on current data

## Hindcasting

Sensitivity to climate fluctuations?

@@

* Comparison to archaeological occurrences

@@

# Discussion

* General trends:
  * Quantified sensitivity of plant ranges to climate change
  * Crop progenitors saw range contractions just before the onset of agriculture? (Moreso than other wild resources??)
  * How could are reconstructed ranges at predicting archaeological assemblages? (+Implications)
* Interesting individual case studies:
  * Wheat progenitors
  * ???

@@

# Conclusion

# References {.appendix}
